Analysis

### 2022-07-02 manuscript/manuscript.Rmd
Final rewriting incorporating CTCF threshold

CTCF binding specificities in the form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases. The JASPAR 2022 [@Castro-Mondragon:2022uv] database lists one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table 1`). The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure 1A`). Although human and mouse CTCF motifs are considered highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM that is expectedly similar to human. SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs. The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the CTCF PWM identified in Jolma et al. study [@Jolma:2013vd]. The Jolma et al. study analyzed the sequence-specific binding of human TFs using high-throughput SELEX and ChIP-sequencing and described 239 distinctly different binding specificities, including a 17bp long CTCF motif dubbed CTCF_full [@Jolma:2013vd]. The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al. [@Schmidt:2012uk], the Ren_20 motif identified by Kim et al. [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al. [@Xie:2007wm] (`Figure 1B`). The CIS-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Table 1`, `Supplementary Figure 1`). The number of human-specific CTCF PWMs likely reflects the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure. These PWMs were used to scan the human and mouse genomes for CTCF binding motifs.

@@

We used the FIMO tool from the MEME suite [@Bailey:2009aa] as one of the oldest and most well-known tool for motif scanning. The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj]. The default p-value cutoff used to generate the package's data is 1E-4. However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 2A`, `Supplementary Figure 2A`), suggesting most CTCF motifs have a relatively poor match. Filtering results at 1E-6 allows focusing on more reliably detected CTCF sites. Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Supplementary Table S1`). Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported 20,000-50,000 sites. Although we provide FIMO results as-is, we recommend filtering and reducing them.

```{r}
fileNameIn1 <- file.path(dir_data, "Figures", "Figure_2.png")
knitr::include_graphics(fileNameIn1)
```
**Figure 2. Properties of CTCF motifs detected by FIMO.** (A) Density plot of p-value distributions of motifs detected using different motif databases in the hg38 genome assembly. Red dashed line - 1E-6 p-value cutoff. (B) Jaccard overlaps among CTCF binding sites detected in the original and liftOver genome assemblies. CTCF sites were detected using JASPAR 2022 MA0139.1 PWM. The correlogram was clustered using Euclidean distance and Ward.D clustering method. White-red gradient indicate low-to-high Jaccard overlaps. Jaccard values are shown in the corresponding cells.

Several databases and papers provide predicted and experimental CTCF binding information. The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatory element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq] (`Table 2`). The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%). The mouse set contained 82,777 records (median size 284bp) similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites. A limitation of this data is that it lacks strand information limiting the opportunity to use this data for CTCF orientation-specific analysis. We added human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.

The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes. The database has not been updated since 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`). As coordinate conversion between genome assemblies (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts. Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure 2B`). liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Supplementary Table S2`). The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Supplementary Figure 2B`).  These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly. Using liftOver, we provide hg19/hg38 and mm9/mm10 versions of predicted CTCF sites from the CTCFBSDB database.

@@

>>>>

CTCF sequence motifs in the form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases. The JASPAR 2022 [@Castro-Mondragon:2022uv] database lists one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table 1`). The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure S1A`). Although human and mouse CTCF motifs are considered to be highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM. SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs. The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the CTCF PWM identified in Jolma et al. study [@Jolma:2013vd]. The Jolma et al. study analyzed the sequence-specific binding of human TFs using high-throughput SELEX and ChIP-sequencing and described 239 distinctly different binding specificities, including a 17bp long CTCF motif dubbed CTCF_full [@Jolma:2013vd]. The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al. [@Schmidt:2012uk], the Ren_20 motif identified by Kim et al. [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al. [@Xie:2007wm] (`Figure S1B`). The CIS-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Figure S1CD`). The number of human-specific CTCF PWMs likely reflects the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure. These PWMs were used to scan the human and mouse genomes for CTCF binding motifs using the FIMO tool from the MEME suite [@Bailey:2009aa] (`Table S1`).

In addition to sequence-derived motifs, several databases provide predicted and experimental CTCF binding sites. The ENCODE Phase III has released SCREEN, a database of cis-regulatory elements (cCREs) for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq] (`Table 2`). The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%). The mouse set contained 82,777 records (median size 284bp) similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites. A limitation of this data is that it lacks strand information limiting the opportunity to use this data for CTCF orientation-specific analysis. We included human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.

The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding sites for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes. The database has not been updated since 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`). As coordinate conversion between genome assemblies (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts. Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure 1A`). liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table S2`). The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Figure S2A`).  These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly. Using liftOver, we provide hg19/hg38 and mm9/mm10 versions of predicted CTCF sites from the CTCFBSDB database.

@@

@@

We used the FIMO tool from the MEME suite [@Bailey:2009aa] as one of the oldest and most well-known tool for motif scanning. The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj]. The default p-value cutoff is 1E-4. However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 1B`, `Figure S2B`), suggesting most CTCF motifs have a relatively poor match. Using the experimental SCREEN CTCF sites as ground truth, we quantified the proportion of true and false positives depending on the FIMO p-value cutoff. We found 1E-6 threshold as the most optimal, providing $\sim80\%$ of true positive CTCF motifs. Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S1`). Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported 20,000-50,000 sites. Although we provide FIMO results using the default settings, for downstream analyses we recommend filtering and reducing them.
